### 2024-04-05 doc/15yrsChange.Rmd
update & automate figure numbering
@llmorreale , FYI

Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure 3b). The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens. Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first two census intervals. However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^. Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval. Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure 4). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.

@@

Consistent with hypothesis 2, biomass losses to mortality outpaced biomass gains from tree growth over the 2018-23 census interval (Figure 3a). For the first time since establishment of the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased between 2018 and 2023 (Figure 2), declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^. Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a). In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023. Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.

@@

 Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure 3c). Plot-wide AWG decreased by `r round((awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013,1) * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval. Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the portion with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^). Woody recruitment flux (AWR), constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d). AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.

@@

Our results reveal that nuisance species have compromised biomass stocks and regenerative capacity at the SCBI ForestGEO plot, reducing the recent and future carbon sink of this forest.
Canopy tree mortality and associated AWM increased, particularly in plot areas affected most heavily by non-indigenous insects and pathogens (hypothesis 1, Figs. 2-3).
These increases in AWM were not offset by biomass increases through AWP, which decreased, or recruitment, resulting in net biomass loss (hypothesis 2, Figs. 2,4).
This net AGB loss occurred despite localized increases in AGB across most of the plot, particularly in areas with low canopy vulnerability (Fig. 6).
Recruitment of canopy tree species was low throughout the plot (Fig. 5), and there was low biomass of canopy species in the understory, particularly in areas with high deer density (Fig. 6). 
Moreover, low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.
Thus, nuisance species have severely compromised both recent and future carbon sequestration in this forest, contradicting models projecting increases in AGB [@wu_uncertainty_2023].

>>>>

Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure `r loss_gain_net`b). The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens. Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first two census intervals. However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^. Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval. Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure `r mortality_by_genera`). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.

@@

Consistent with hypothesis 2, biomass losses to mortality outpaced biomass gains from tree growth over the 2018-23 census interval (Figure `r loss_gain_net`a). For the first time since establishment of the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased between 2018 and 2023 (Figure `r biomass_trends`), declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^. Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure `r loss_gain_net`a). In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023. Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.

@@

 Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure `r loss_gain_net`c). Plot-wide AWG decreased by `r round((awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013,1) * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval. Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the portion with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^). Woody recruitment flux (AWR), constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure `r loss_gain_net`d). AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.

@@

Our results reveal that nuisance species have compromised biomass stocks and regenerative capacity at the SCBI ForestGEO plot, reducing the recent and future carbon sink of this forest.
Canopy tree mortality and associated AWM increased, particularly in plot areas affected most heavily by non-indigenous insects and pathogens (hypothesis 1, Figs. `r loss_gain_net`- `r mortality_by_genera`).
These increases in AWM were not offset by biomass increases through AWP, which decreased, or recruitment, resulting in net biomass loss (hypothesis 2, Figs. `r loss_gain_net`, `r biomass_trends`).
This net AGB loss occurred despite localized increases in AGB across most of the plot, particularly in areas with low canopy vulnerability (Fig. `r regen_failure`).
Recruitment of canopy tree species was low throughout the plot (Fig. `r recruitment_by_genera`), and there was low biomass of canopy species in the understory, particularly in areas with high deer density (Fig. `r regen_failure`). 
Moreover, low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.
Thus, nuisance species have severely compromised both recent and future carbon sequestration in this forest, contradicting models projecting increases in AGB [@wu_uncertainty_2023].
